Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-01-26 cabios-template2/bigsnpr-review.tex
before mike review

\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

@@

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. {\color{red} This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.} \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

{\color{red}
Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, it has been required to continuously optimize algorithms and this has led to the reimplementation of some software that would have become obsolete otherwise. For computing Principal Component Analysis, commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which is a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including R packages snpStats, GenABEL, SNPRelate and GWASTools. Packages snpStats and GenABEL, which have been widely used and cited, are now outdated with current data sizes -- package snpStats still uses in-memory data and package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
Solving the memory issues for languages such as R would give access to a broad range of tools for data analysis that have already been implemented. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
}

@@

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
For instance, one might like using R because R makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores. We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.

@@

We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008} and of the R package BEDMatrix\footnote{https://github.com/QuantGen/BEDMatrix} which provides memory-mapping directly for binary PLINK files. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso and RSpectra \cite[]{RSpectra2016,Zeng2017}.

Memory-mapping provides transparent and faster access than standard read/write operations. 
When an element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. 
When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. 
All this is managed by the Operating System so that it is seamless and efficient. 
It means that if the same chunks of data are used repeatedly, it will be very fast the second time they are accessed, the third time and so on. 
Of course, if the memory size of the computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

{\color{red}
We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values). So, we use a single data format that can store both genotype calls and dosages.

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}). While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. Package bigstatsr uses the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.

\begin{figure}[!htpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
\end{figure}

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the MAF as the statistic of importance. 
Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest minor allele frequency (MAF) is kept.
Yet, in some worst-case scenario, this pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, we tested the effect of using clumping with the MAF instead of pruning for the thinning step before PCA and the results were very similar.
}

As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in long-range LD regions begore computing PCs. 
However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping (with MAFs), then outliers SNPs are detected using the robust Mahalanobis distance as implemented in method pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD of package bigsnpr and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
{\color{red}
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
}
%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).

The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
{\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Euesden2015}. 
On the contrary, the second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables \textit{at once}, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are very fast sparse linear and logistic regressions. These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of the choice of the regularization parameter.

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Materials. 
{\color{red}
In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
}

@@

We present the results for four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. For the step concerning the computation of 10 principal components, we further compare methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (for genotyped SNPs only). 
We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly, requires only 20 lines of R code and there is no need to write temporary files or objects because functions of packages bigstatsr and bigsnpr have parameters which enable subsetting of the genotype matrix without having to copy it.

\begin{figure}[!htpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}

>>>>

\title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

@@

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. {\color{red} This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.} \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement fast and accurate computations of principal component analysis and association studies, functions to remove SNPs in linkage disequilibrium and algorithms to learn polygenic risk scores on millions of SNPs. We illustrate applications of the two R packages by analyzing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary materials are available at \textit{Bioinformatics}
online.}

@@

{\color{red}
Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, it has been required to continuously optimize algorithms and this has led to the re-implementation of some software that would have become obsolete otherwise. For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}. Package GenABEL, which has been widely used and cited, does not handle standard binary PLINK files and has not been updated since 2013.
Solving the memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis. Fortunately, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
}

@@

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
For instance, one might like using R because this programming language makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice. We report execution times along with the code to perform major computational tasks.

@@

We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008} and of the R package BEDMatrix\footnote{https://github.com/QuantGen/BEDMatrix} which provides memory-mapping directly for binary PLINK files. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as if it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso and RSpectra \cite[]{RSpectra2016,Zeng2017}.

Memory-mapping provides transparent and faster access than standard read/write operations. 
When an element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. 
When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. 
All this is managed by the operating system so that it is seamless and efficient. 
It means that if the same chunks of data are used repeatedly, it will be very fast the second time they are accessed, the third time and so on. 
Of course, if the memory size of the computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

{\color{red}
We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored in one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed'' which can store only genotype calls. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values). So, we use a single data format that can store both genotype calls and dosages.

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition and the other one based on randomized projections, respectively named big\_SVD and big\_randomSVD (Figure~\ref{fig:svd}). While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. Package bigstatsr uses the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
\end{figure}

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest MAF is kept.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, we tested the effect of using clumping with the MAF instead of pruning for the thinning step before PCA and the results were very similar.
}

As mentioned above, the third approach that is generally combined with pruning and consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in these regions before computing PCs. 
However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping (with MAFs), then outliers SNPs are detected using the robust Mahalanobis distance as implemented in method pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD of package bigsnpr and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\left(\frac{\hat{p}}{1-\hat{p}}\right)} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
{\color{red}
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
}
%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).

The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
{\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Euesden2015}. 
On the contrary, the second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables \textit{at once}, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are very fast sparse linear and logistic regressions. These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of choosing the regularization parameter.

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the supplementary materials. 
{\color{red}
In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
}

@@

We present the results for four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. For the step concerning the computation of 10 principal components, we further compare methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how SNPs are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly, requires only 20 lines of R code and there is no need to write temporary files or objects because functions of packages bigstatsr and bigsnpr have parameters which enable subsetting of the genotype matrix without having to copy it.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}
